Stage-Splitting a Modal Language

### 2014-10-29 paper/splitting.tex
worked on global structure of splitting section

The goal of splitting is to take a multistage term and produce two new terms:
one with all of the stage \bbone\ content, and one with all of the stage \bbtwo\ content.
Exactly how this goal manifests depends on the form of the input. 
We consider four cases: stage \bbone\ values, stage \bbone\ terms, stage \bbtwo\ values,
and stage \bbtwo\ terms.

We cover stage \bbtwo\ value splitting first, because it is trivial by construction.
Since there is no stage \bbtwo\ expression that both is a value
and contains \bbone\ subexpressions, 
we say that stage \bbtwo\ values must contain no stage \bbone\ content.
Accordingly, there is nothing really to split.

\subsection{Partial Values}

Partial values, along with future environments, are produced by partially
evaluating stage \bbone\ terms. Although partial values have no more stage
\bbone\ computation remaining, they may contain variables bound by the future
environment, and thus, may require stage \bbtwo\ computation.

Splitting a partial value $v$ creates two single-stage values $v_1$ and $v_2$,
where $v_w$ contains only the information required for stage $w$ evaluation of
$v$. Splitting partial values is implemented by the judgment $v \vsplito
[v_1,v_2]$ defined in \ref{fig:valSplit}. 
At product, sum, and recursive types, splitting acts recursively, gathering all
subterms' stage $w$ content into $v_w$. At base types, splitting assigns $v$
entirely to stage \bbone\ (and \texttt{()} to stage \bbtwo); on the other hand,
at $\fut$ types, splitting assigns $v$ (which must have the form $\next~\hat y$)
entirely to stage \bbtwo\ (and \texttt{()} to stage \bbone). This process of
exposing only the relevant structure of a term is called \emph{masking}.

\ur{The case for ``fold'' seems missing in partial value splitting rules.}

@@

\subsection{General Splitting}

Splitting arbitrary terms in \lang\ again results in two terms containing only
the stage \bbone\ or stage \bbtwo\ content of the original term; these terms are
called, respectively, the {\em precomputation} and {\em resumer}.
%
For example, the stage \bbtwo\ term
\begin{lstlisting}
2`hold{`1` 1+2 `2`} < hold{`1` 3+4 `2`} - 5`
\end{lstlisting}
splits into a precomputation \verb|(1+2,3+4)| and a resumer
\verb|fn (L1,L2) => L1 < L2-5|, with the property that the resumer applied to
the precomputation must agree with the original term. This process of hoisting
all the stage \bbone\ subterms out of a term is called \emph{lifting}.

@@

\subsection {Formal Setup}

%in rule-for-rule correspondence with the typing rules $@\bbone$ and $@\bbtwo$.
Splitting is implemented by the judgments $\splitonesym$ and $\splittwosym$,
which take stage \bbone\ and stage \bbtwo\ terms respectively; these judgments
are defined in \cref{fig:termSplitOne,fig:termSplitTwo} inductively on the
structure of the term. Both judgments produce two terms in an unstaged language,
\langmono, whose grammar is given in \ref{fig:monoGrammar}. This language is
untyped and contains no staging features, but otherwise has the same term
constructors as \lang.
%\footnote{Since \langmono\ is untyped, we could encode everything with just functions.}.

The $\splitonesym$ judgment sends a stage \bbone\ term $e$ to a combined term
$c$ and resumer $l.r$, while $\splittwosym$ sends a stage \bbtwo\ term $e$ to a
precomputation $p$ and a resumer $l.r$. In both cases, the term $e$ is open on
variables in the context $\Gamma$. We write the rules using pattern notation
when appropriate.
%For concision, we represent resumers as a term open on a single variable,
%rather than as functions.

\ur{Reformat rules to make them fit into the paper.}

The $\splittwosym$ rules are conceptually simple and almost all very similar.  
As mentioned before, the only responsibility of $\splittwosym$ is to lift out the precomputation
from contained stage \bbone\ code into a single expression $p$, 
leaving a resumer $l.r$ to interpret the results of that precomputation.
Since base types and variables have no stage \bbone\ content, their precomputation is trivial;
that is, $p$ is a unit value and $l$ is never used in $r$.
At every other expression except $\prev$, the $\splittwosym$ rule works by bundling 
the precomputations of the constituent parts, and then unbundling them with a pattern,
with the resumer precisely resembling the original expression.

@@

@@

\paragraph {Expression Splitting at Values}

When applied to partial values, term splitting always produces a trivial
precomputation, in the sense that the output is a nested sequence, possibly
empty, of tuples and injections of \texttt{()}. The intuition is that a partial
value has no remaining stage \bbone\ computation, and hence no precomputation.

@@

%\subsection {Boundary Type Worst Case}
%
%\TODO add the example which is the worst case for figuring out boundary types
%

>>>>

The goal of splitting is to take a multistage term and produce two new terms:
one with all of the stage \bbone\ content, and one with all of the stage \bbtwo\ content.
Exactly what content is implied by this goal depends on the form of the input. 
There are four cases: stage \bbone\ partial values, stage \bbone\ terms, stage \bbtwo\ values,
and stage \bbtwo\ terms.

In all cases, the outputs of splitting are terms in an unstaged language, \langmono. 
This language is untyped and contains no staging features, 
but otherwise has the same term constructors as \lang. 
It has a evaluation judgment $\redsym$, which works in the expected way to produce unstaged values.

\subsection {Stage \bbtwo\ Terms}

We cover stage \bbtwo\ term splitting first, because it has the most familiar signature.
In spirit, splitting takes a term $\coltwo e A$ and produces from it a first stage term
$p : \tau$ (called the {\em precomputation}) and a second stage term $r : \tau \to A$ (called the {\em resumer}).
Here, the responsibility of $p$ is to produce an intermediate data structure, 
and the responsibility of $r$ is to consume that data structure and produce the same final value as $e$.

Our actual presentation of stage splitting differs from this ideal in two ways.
Firstly, the outputs are untyped, for reasons discussed later.
Secondly, we do not actually represent the resumer as a function 
but instead as a term open on a variable, typically $l.r$.
This helps keep the splitting rules concise.

 For example, the stage \bbtwo\ term
\begin{lstlisting}
2`hold{`1` 1+2 `2`} < hold{`1` 3+4 `2`} - 5`
\end{lstlisting}
splits into a precomputation \verb|(1+2,3+4)| and a resumer
\verb|fn (L1,L2) => L1 < L2-5|, with the property that the resumer applied to
the precomputation must agree with the original term. This process of hoisting
all the stage \bbone\ subterms out of a term is called \emph{lifting}.

Splitting is implemented by the judgment $\splittwosym$,
the rules for which are given in \cref{fig:termSplitTwo}.  
It operates inductively on the structure of the term.  
As mentioned before, the only responsibility of $\splittwosym$ is to lift out the precomputation
from contained stage \bbone\ code into a single expression $p$, 
leaving a resumer $l.r$ to interpret the results of that precomputation.
Since base types and variables have no stage \bbone\ content, their precomputation is trivial;
that is, $p$ is a unit value and $l$ is never used in $r$.
At every other expression except $\prev$, the $\splittwosym$ rule works by bundling 
the precomputations of the constituent parts, and then unbundling them with a pattern,
with the resumer precisely resembling the original expression.
The $\prev$ rule depends on splitting stage \bbone\ terms, which is covered later.

This is enough to define a correctness goal for splitting closed terms.

\paragraph {Correctness of stage \bbtwo\ splitting.}
If $e : A \splittwosym [p,l.r]$, then $e \equiv (\letin l p r)$

\subsection {Stage \bbtwo\ Values}

We next consider splitting for the special case of stage \bbtwo\ values.
Observing that no stage \bbtwo\ expression both is a value and contains stage \bbone\ subexpressions, 
we say that stage \bbtwo\ values must contain no stage \bbone\ content.
That is, stage \bbtwo\ value splitting is trivial by construction.

When applied to partial values, stage \bbtwo\ term splitting always produces a trivial
precomputation, in the sense that the output is a nested sequence, possibly
empty, of tuples and injections of \texttt{()}. The intuition is that a partial
value has no remaining stage \bbone\ computation, and hence no precomputation.

\subsection{Stage \bbone\ Partial Values}

Partial values, along with future environments, are produced by partially evaluating stage \bbone\ terms. 
They, unlike stage \bbtwo\ values, {\em can} have content at both stages.
As discussed above, the {\tt quickselect} function itself is an example of a multi-stage values.

Although partial values have no more stage
\bbone\ computation remaining, they may contain $\next$ blocks
(the contents of which must be variables bound by the future environment).
The goal of splitting for stage \bbone\ partial values is therefore 
to separate the contents of $\next$ blocks from the rest of the value, while maintaining some structure.

Splitting a partial value $v$ creates two unstaged values $v_1$ and $v_2$,
where $v_w$ contains only the information required for stage $w$ evaluation of
$v$. Splitting partial values is implemented by the judgment $v \vsplito
[v_1,v_2]$ defined in \ref{fig:valSplit}. 
At product, sum, and recursive types, splitting acts recursively, gathering all
subterms' stage $w$ content into $v_w$. At base types, splitting assigns $v$
entirely to stage \bbone\ (and \texttt{()} to stage \bbtwo); on the other hand,
at $\fut$ types, splitting assigns $v$ (which must have the form $\next~\hat y$)
entirely to stage \bbtwo\ (and \texttt{()} to stage \bbone). This process of
exposing only the relevant structure of a term is called \emph{masking}.

\ur{The case for ``roll'' seems missing in partial value splitting rules.}

@@

\subsection{Stage \bbone\ Term Splitting}

@@

Splitting is implemented by the judgment $\splitonesym$,
the rules for which are given in \cref{fig:termSplitOne}.  
It operates inductively on the structure of the term.  
The $\splitonesym$ judgment sends a stage \bbone\ term $e$ to a combined term
$c$ and resumer $l.r$.

@@

\subsection {Notable Properties}

\nr{These points will probably get lifted into the preceding sections.}

@@

@@

\subsection {Boundary Type Worst Case}

\TODO add the example which shows that we need dynamic types
